A Java Framework For Embarrassingly Parallel Problems

### 2013-12-13 paper/paper.tex
done paper

%TODO
\begin{figure}
    \begin{center}
        \includegraphics{figures/ep.jpg}
    \end{center}
    \caption{Embarrassingly Parallel Model}
    \floatfoot{http://www.coopsoft.com/ar/i/fjGraph.jpg}
    \label{fig:ep}
\end{figure}

@@

Ka/Ks Calculator is not the only embarrassingly parallel program that has the
potential to take several days to run on a data set. In fact, there are several
embarrassingly parallel programs in the biological sciences. Basic Local
Alignment Search Tool, or BLAST\footnote{blast.ncbi.nlm.nih.gov}, is an
algorithm that compares many aspects of DNA sequences, like nucleotides or amino
acid sequences. In addition, the user can also compare the sequences to a
database of sequences to identity known subservience's. There are many
implementations of BLAST, some of them already have parallel versions.  Hovever,
many applications do not have parallel processing capabilities. There exists a
class of problems that organize large amounts of data into similar groups called
clusters. Two applications, octupus\footnote{octupus.sourceforge.com} and
USEARCH\footnote{https://www.drive5.com/usearch}, are examples of programs in
this class that do not have parallel versions. With several applications that
would benefit from parallelism in this field, it was obvious it would be
desirable to develop a framework for the parallelization task.

@@

Amazon also offers another service called Amazon Elastic
MapReduce\footnote{https://aws.amazon.com/elasticmapreduce}. This combines its
virtual instance framework with Apache
Hadoop\footnote{https://hadoop.apache.org}. As seen in figure \ref{fig:map},
Hadoop breaks down the input data into parts, runs them in parallel with the
'Map' function and then combines the results with the 'Reduce' function.
DryadLINQ\footnote{https://research.microsoft.com/apps/mobile/showpage.aspx?page=/en-us/projects/draydlinq}
is a framework developed by Microsoft Research that allows the end-user to
create directed acyclic computation graphs that run on Windows High Performance
Computer(HPC)\footnote{https://www.microsoft.com/hpc} clusters.  Gunarathne et
al. compared the services using several applications to analyze performance,
cost and usability.

@@

The only requirements for a program to be wrapped by our framework are that it
processes the input in an embarrassingly parallel manner and that the input comes
from a file. The framework, shown in figure \ref{fig:uml}, supports two forms of input: a directory of files that each contain
one unit of work, which is represented by the Chunk class, or a single file that
the framework can break up into Chunks. For example, the Ka/Ks Calculator's
input consists of a header and then two protein sequence lines followed by a
blank line. The framework uses that blank line to know when one Chunk of work
has ended and another has begun. The user can input a regular expression to
define when lines of input are separated into Chunks. This occurs in the
ChunkManager class.

The other purpose of the ChunkManager class is to combine the resulting Chunks
back together when the worker threads have finished. Once the work has been
done, all of the separate results are spread across several files. Those results
must be retrieved and combined to construct the final output file. CONTINUED

This output
file must look the same as the output file would look if the program were run
serially; the output may need to come in the same order as the inputs in the
input file and headers may need to be deleted from the separate files and added
to the combined file. End-users can also choose to write their own merge code in
Java, using the CustomMerge class, or write their own independent merge script.

REWRITE ABOVE

@@

We evaluated the convenience and performance of the framework on two different
applications: the Ka/Ks Calculator and a sequence clustering algorithms,
ocutupus. In all cases, we compared the framework based application with the
original using different data partitions and different thread counts.

\subsection{Ka/Ks Calculator}
Ka/Ks Calculator was run through the framework with five different samples, each
containing sixteen units of work. Each sample was run by the framework thread
counts, starting at 1 and increasing by multiples of two until the runtimes
leveled off. The samples were then all run again at 100\% processor usage (16
processors). These tests were run with minimal other activity on the server.

@@

Figure \ref{fig:time} summarized the results; each line shows the time taken for
each chunk of the data. As the numbers of threads increases, the runtime of the
program decreases exponentially. In a few cases, there was an increase in time
when attempting to max out the CPU usage. This occurs because the framework has
to wait for the CPU to be finished with its previous work before it can start
any further jobs.

Figure \ref{fig:speedup} shows the speedup of the Ka/Ks Calculator program. 
In most of the samples, the speedup seems to level off between 6 to 10 workers,
and even goes down when increased to 16 workers. Running the framework at 50\% 
of the CPU is more efficient than running at 100\% of the CPU in this 
situation. The decrease in speedup from 8 workers to 16 workers shows this.
This is not true for all cases, as seen in sample E. This sample has a higher
speedup at 16 workers because there are multiple pieces of work toward the end
that take a majority of the runtime. With 16 workers, these chunks were started
immediately instead of later during the total runtime of the framework.

@@

A version of octupus was also run through the framework and was tested against
two different sized data sets. Both sets were first clustered at 90\%, 95\% and
97\% minimum similarity. The runtime for those tests can be seen in table
\ref{tab:octu1}. The resulting clusters from the 90\% similarity run were used as
the input for the framework's runs, which were all set for 97\% similarity. 
The framework in this case was run at six, eight and ten workers. Table 
\ref{tab:octu2} shows that it is slightly faster to cluster first at 90\% and 
then run the resulting clusters through the framework at a higher similarity.

\begin{table}
    \begin{center}
    \begin{tabular}{|l|l|l|}
    \hline
        Number of Sequences & Minimum Similarity & Time(s) \\ \hline
        1500                & 90\%               & 99      \\
        ~                   & 95\%               & 115     \\
        ~                   & 97\%               & 131     \\ \hline
        6500                & 90\%               & 425     \\
        ~                   & 95\%               & 496     \\
        ~                   & 97\%               & 603     \\ \hline
    \end{tabular}
    \caption{Times from serial octupus at various similarity levels}
    \end{center}
    \label{tab:octu1}
\end{table}

\begin{table}
    \begin{center}
    \begin{tabular}{|l|l|l|}
        \hline
        Number of Sequences & Number of Workers & Time(s) \\ \hline
        1500                & 6                 & 34      \\
        ~                   & 8                 & 27      \\
        ~                   & 10                & 27      \\ \hline
        6500                & 6                 & 127     \\
        ~                   & 8                 & 127     \\
        ~                   & 10                & 123     \\ \hline
    \end{tabular}
    \caption{Times from octupus through the framework at 97\% similarirty}
    \end{center}
    \label{tab:octu2}
\end{table}

@@

It is clear that the framework provides an effective yet simple to use module to
add parallelism to a specific class of file-oriented embarrassingly parallel
problems. Speedup of 4x can be achieved on simple samples with the same number
of chunks as workers, and possibly slightly higher with real data, as seen with
the octupus data. While many consumer-level computers will not have many options
when it comes to chosing the proccesing power, the framework will help
universities and companies who have the raw computing power, and are just not
using it to their full potential.

>>>>

\begin{figure}
    \begin{center}
        \includegraphics{figures/ep.jpg}
    \end{center}
    \caption{Embarrassingly Parallel Model}
    \floatfoot{http://www.coopsoft.com/ar/i/fjGraph.jpg}
    \label{fig:ep}
\end{figure}

@@

Ka/Ks Calculator is not the only embarrassingly parallel program that has the
potential to take several days to run on a data set. In fact, there are several
embarrassingly parallel programs in the biological sciences. Basic Local
Alignment Search Tool, or BLAST\footnote{blast.ncbi.nlm.nih.gov}, is an
algorithm that compares many aspects of DNA sequences, like nucleotides or amino
acid sequences. In addition, the user can also compare the sequences to a
database of sequences to identity known subservience's. There are many
implementations of BLAST, some of them already have parallel versions.  However,
many applications do not have parallel processing capabilities. There exists a
class of problems that organize large amounts of data into similar groups called
clusters. Two applications, octupus\footnote{octupus.sourceforge.com} and
USEARCH\footnote{https://www.drive5.com/usearch}, are examples of programs in
this class that do not have parallel versions. With several applications that
would benefit from parallelism in this field, it was obvious it would be
desirable to develop a framework for the parallelization task.

@@

Amazon also offers another service called Amazon Elastic
MapReduce\footnote{https://aws.amazon.com/elasticmapreduce}. This combines its
virtual instance framework with Apache
Hadoop\footnote{https://hadoop.apache.org}. As seen in Figure \ref{fig:map},
Hadoop breaks down the input data into parts, runs them in parallel with the
'Map' function and then combines the results with the 'Reduce' function.
DryadLINQ\footnote{https://research.microsoft.com/apps/mobile/showpage.aspx?page=/en-us/projects/draydlinq}
is a framework developed by Microsoft Research that allows the end-user to
create directed acyclic computation graphs that run on Windows High Performance
Computer(HPC)\footnote{https://www.microsoft.com/hpc} clusters.  Gunarathne et
al. compared the services using several applications to analyze performance,
cost and usability.

@@

The only requirements for a program to be wrapped by our framework are that it
processes the input in an embarrassingly parallel manner and that the input
comes from a file. The framework, shown in Figure \ref{fig:uml}, supports two
forms of input: a directory of files that each contain one unit of work, which
is represented by the Chunk class, or a single file that the framework can break
up into Chunks. For example, the Ka/Ks Calculator's input consists of a header
and then two protein sequence lines followed by a blank line. The framework uses
that blank line to know when one Chunk of work has ended and another has begun.
The user can input a regular expression to define when lines of input are
separated into Chunks. This occurs in the ChunkManager class.

The other purpose of the ChunkManager class is to combine the resulting Chunks
back together when the worker threads have finished. Once the work has been
done, all of the separate results are spread across several files. Those results
must be retrieved and combined to construct the final output file. This combined
output file looks as if the program were run serially: the output must come in
the same order and the headers need to be added. End-users can also choose to
write their own merge code in Java, using the CustomMerge class, or write their
own independent merge script.

@@

We evaluated the convenience and performance of the framework on two different
applications: the Ka/Ks Calculator and a sequence clustering algorithms,
octupus. In all cases, we compared the framework based application with the
original using different data partitions and different thread counts.

\subsection{Ka/Ks Calculator}

Ka/Ks Calculator was run through the framework with five different samples, each
containing sixteen units of work. Each sample was run with a range of thread
counts, starting at 1 and increasing by a multiple of two until the runtimes
leveled off and then once more with the maximum 16 processors. These tests were
run with minimal other activity on the server.

@@

Figure \ref{fig:time} summarizes the results; each line shows the time taken for
each chunk of the data. As the numbers of threads increases, the runtime of the
program decreases exponentially for a while and then begins to flatten between 6
and 10 processors. In a few cases, there was an increase in time when attempting
to max out the CPU usage. This occurs because the framework has to wait for the
CPU to be finished with its previous work before it can start any further jobs.
Figure \ref{fig:speedup} shows the speedup of the Ka/Ks Calculator program.  In
most of the samples, the speedup seems to level off between 6 to 10 workers, and
even goes down when increased to 16 workers. Running the framework at 50\% of
the CPU is more efficient than running at 100\% of the CPU in this situation.
The decrease in speedup from 8 workers to 16 workers shows this.

This is not true for all cases, as seen in sample E. This sample has a higher
speedup at 16 workers because there are multiple pieces of work toward the end
that take a majority of the runtime. With 16 workers, these chunks were started
immediately instead of later during the total runtime of the framework. Another
random sample was taken to show that sample E was an anomaly. This sample had
similar runtimes at 8 and 16 workers, resulting in a negligible speedup increase
when going up to 100\% of the CPU. This random sample behaving similarly to the
majority of the sample shows that sample E behaved differently due to having a
majority of its work later in the file. Some tests were run in the early stages
of development to observe the behavior of sorting the data by length. In terms
of Ka/Ks, there was no correlation between length of sequences and calculation
time, so there was no way to sort the data to ensure the chunks of work that
take the quickest segments happen last. This would prevent anomalies like sample
E.

@@

A version of octupus was also run through the framework and was tested against
two different sized data sets. Both sets were first clustered at 90\%, 95\% and
97\% minimum similarity with only one processor. The runtime for those tests can
be seen in Table \ref{tab:octu1}. The resulting clusters from the 90\%
similarity run were used as the input for the framework's runs, which were all
set for 97\% similarity.  The framework in this case was run at six, eight and
ten workers. Table \ref{tab:octu2} shows that it is slightly faster to cluster
first at 90\% and then run the resulting clusters through the framework at a
higher similarity.

\begin{table}
    \begin{center}
    \begin{tabular}{|l|l|l|}
    \hline
        Number of Sequences & Minimum Similarity & Time(s) \\ \hline
        1500                & 90\%               & 99      \\
        ~                   & 95\%               & 115     \\
        ~                   & 97\%               & 131     \\ \hline
        6500                & 90\%               & 425     \\
        ~                   & 95\%               & 496     \\
        ~                   & 97\%               & 603     \\ \hline
    \end{tabular}
    \caption{Times from serial octupus at various similarity levels}
    \label{tab:octu1}
    \end{center}
\end{table}

\begin{table}
    \begin{center}
    \begin{tabular}{|l|l|l|}
        \hline
        Number of Sequences & Number of Workers & Time(s) \\ \hline
        1500                & 6                 & 133      \\
        ~                   & 8                 & 126      \\
        ~                   & 10                & 126      \\ \hline
        6500                & 6                 & 552     \\
        ~                   & 8                 & 552     \\
        ~                   & 10                & 548     \\ \hline
    \end{tabular}
    \caption{Times from octupus through the framework at 97\% similarity}
    \label{tab:octu2}
    \end{center}
\end{table}

@@

It is clear that the framework provides an effective yet simple to use module to
add parallelism to a specific class of file-oriented embarrassingly parallel
problems. Speedup of 4x can be achieved on simple samples with the same number
of chunks as workers, and possibly slightly higher with real data, as seen with
the octupus data. While many consumer-level computers will not have many options
when it comes to choosing the processing power, the framework will help
universities and companies who have the raw computing power, and are just not
using it to their full potential.
